Time-variable surface mapping of exoplanets: an application to Io

### 2020-10-08 paper/paper.tex
Updating abstract

\begin{abstract}
Jupiter's moon Io is the most volcanically active body in the Solar System with hundreds of active volcanoes varying in intensity on different timescales.
Existing photometric observations in the near infrared taken during occultations by Jupiter and Galilean moons encode a wealth of information about its surface.
As such, Io is an ideal testbed for models of time-variable exoplanet surfaces.
We build a single probabilistic model of Io's surface using Nonnegative Matrix Factorization to infer one map per light curve.
The model is built on top of the code starry \href{https://rodluger.github.io/starry/}{\color{linkcolor}\faGithub} which can compute occultation light curves and phase curves in both emitted and reflected light. It also enables efficient inference with gradient based samplers.
We use the inferred map to study the time variability of prominent volcanoes over a timescale of decades.
While we mostly focus on Io, the lessons we learn are directly applicable to mapping exoplanets.
We discuss this application towards the end of the paper.\href{https://github.com/fbartolic/volcano}{\color{linkcolor}\faGithub}

@@

\subsection{Schmixels vs. spherical harmonics}
\label{ssec:schmixels_vs_harmonics}
To test inference with the static map model defined in Section~\ref{ssec:static} 
we use \textsf{starry} to generate a single simulated light curve of an occultation of Io by Jupiter from an $l=30$ map.
The simulated map consists of a spherical harmonic expansion of a bright spot with a Gaussian profile which we add using the built in \textsf{add\_spot} function in \textsf{starry}.
The expansion is in the quantity $\cos(\Delta\theta)$ where $\Delta\theta$ is the angular separation between the center of the spot and another point on the surface of the sphere. 
We place the spot at 13 degrees latitude and 51 degrees East longitude, set the diameter of the spot ($2\Delta\theta$) to 5 degrees and the amplitude to 55\% of the total luminosity of a spotless map.
We take ephemeris parameters (assumed to be fixed) from a real occultation and assume an observational cadence of 1s and errorbars equal to 2\% of the total flux, similar to high quality IRTF light curves.
We use this dataset to test the difference between fitting a model in the spherical harmonic basis (Eq.~\ref{eq:linear_model}) and in the schmixel basis (Eq.~\ref{eq:linear_model_pix})
by fitting for an $l=20$ map.

@@

The schmixel model with Half Gaussian priors produces a more sensible map in which the unobserved hemisphere has much lower intensity but the arc is less sharply defined and more extended and as a result the predicted flux does not rise steeply enough compared to the light curve.
The map from a model with an Exponential prior on the schmixels provides the best fit.
It is the sparsest of the three maps, having a single bright spot at the center of the projected limb of Jupiter at the location of the true spot with extended tails along the limb.
In principle, the sparsity is a consequence of the Exponential prior which has significant probability mass near zero, meaning that it pushes most schmixels towards zero intensity, and a heavy tail towards large intensities.
It favors parse maps with near zero intensity everywhere except for a few high intensity regions which is exactly the kind of prior information we want to assume for Io because we expect spot like features on its surface.
In practice, as shown in Figure~\ref{fig:schmixels_to_pixels}, the Exponential prior 
implies a more Gaussian like prior in spherical harmonic space and yet it still gives noticeably different looking map than a Gaussian (NOT CONVINCING).

%Figure~\ref{fig:pixels_hist} shows the cumulative distributions of pixels and schmixels corresponding to the three inferred maps shown in Figure~\ref{fig:schmixels_vs_harmonics}, shown in the same order as the inferred maps.
%The plot on the left shows the distribution of pixels in the true map (blue line) 
%and the inferred pixels $\mathbf{P}\,\mathbf{y}$ (orange).
%The true distribution rapidly increases towards unity because the majority of pixels have near zero intensity. 
%There is a very small tail towards negative values which is due to the fact that we generate the true map at $l=30$ and even at that resolution there is still some ringing present leading to negative intensities.
%The inferred distribution has a very different shape with approximately 20\% of pixels having negative values and a much more gradual increase towards unity.
%
%The second subplot shows the distribution of both pixels (orange) and schmixels (green) corresponding to the model with Half Gaussian priors on the schmixels.
%The distribution of pixels and schmixels in this case is nearly identical meaning that the approximate mapping $\mathbf{P}^\dagger$ is good enough.
%In the case of an Exponential prior on schmixels (right) the shape of the schmixel and pixel distributions is similar to the Gaussian prior solution except for a small discrepancy between the two distributions.
%The schmixel distribution has no mass at negative values (by definition), but the inferred pixel distribution has a very small negative tail.
%The discrepancy is likely due to the fact that the Exponential prior favors more concentrated regions of high intensity which induce more ringing artefacts with some negative pixels.
%
%\begin{figure}[h!]
%    \begin{centering}
%        \includegraphics[width=\linewidth]{figures/pixels_hist.pdf}
%    \oscaption{pixels_hist}{%
%        Cumulative distribution of pixels in the true map and the and schmixels/pixels of inferred maps shown in Fig.~\ref{fig:schmixels_vs_harmonics}.
%        The distribution of pixels in the inferred map from the $Y_{lm}$ model (left) is very different from the truth and it has a significant left tail towards negative values.
%        On the other hand, the distribution of pixels in the schmixel model (middle and right) is much closer to the truth and it has a negligible fraction of negative intensities.
%        The distribution of pixels closely follows the distribution of schmixels for both choices of priors on schmixels. 
%        \label{fig:pixels_hist}
%    }
%    \end{centering}
%\end{figure}

\subsection{Smoothing out spurious features}
\label{ssec:spurious_features}
To suppress ringing which results in regions of negative intensity surrounding a spot like feature as in Figure~\ref{fig:schmixels_vs_harmonics}, we choose to apply a spatial smoothing filter to the spherical harmonics.
Mathematically, the filtering operation is a convolution between the map and some kernel function $B(\theta,\phi)$.
Assuming both the map and the kernel function are expanded in terms of spherical harmonics, the convolution operation is simply a multiplication between the two sets of spherical harmonic coefficients.
We choose a simple azimuthally symmetric Gaussian kernel function given by
\begin{equation}
    B(\theta)=\frac{1}{\sqrt{2 \pi \sigma_s^{2}}}\exp \left(-\theta^{2} / 2 \sigma_s^{2}\right)
\end{equation}
where $\sigma_s$ sets the characteristic scale of the smoothing.
This function can be expended in terms of spherical harmonics as
\begin{equation}
    B(\theta)=\sum_{l=0}^{\infty}\left(\frac{2 l+1}{4 \pi}\right) B_{l} \,\mathcal{P}_{l}(\cos \theta)
\end{equation}
where $B_l$ are the spherical harmonic coefficients and $\mathcal{P}_l$ are the associated Legendre polynomials.
They depend only on $l$ because all nonzero $m$ modes vanish due to azimuthal symmetry.
For $\sigma_s\ll 1$ $B_l$ can be approximated as \citep{seon2007,white1995}
\begin{equation}
    B_l\simeq \exp\left[-\frac{1}{2}l(l+1)\sigma_s^2\right]
\end{equation}
The Gaussian filter results in an exponential cutoff for scales smaller than $l\sim \sigma_s^{-1}$.

In Figure~\ref{fig:smoothing_kernel} we show the effect of (Gaussian) smoothing on a spherical harmonic expansion of a spot with a Gaussian intensity profile.
The expansion is in the quantity $\cos\Delta\theta$ where $\Delta\theta$ is the angular separation between the center of the spot and another point on the surface of the sphere.
We place the spot at zero degrees latitude and longitude and set the size of the spot $\Delta\theta$ to 5 degrees.
All three subplots show the exact profile of the spot (black line) and expansions up to three different orders $l$ (colored lines).
The leftmost subplot shows the expansion with no smoothing ($\sigma_s=0$) in which case the symmetric ringing around the center of the spot is clearly visible even at relatively high order ($l=20$).
The middle subplot shows an intermediate level of smoothing where $\sigma_s=0.1$, meaning that all features on scales above $l\approx 10$ are exponentially suppressed.
In this case the negative ringing is a lot less visible albeit at the cost of having a slightly larger spot because suppressing higher order harmonics necessarily means that we lose some ability to represent smaller scale features.
This is most apparent in the rightmost subplot where $\sigma_s=0.2$. 
Although there is practically no ringing, expansions at $l=10$, $l=15$ and $l=20$ look almost the same because all coefficients above $l=5$ are significantly suppressed.

\begin{figure}[h!]
    \begin{centering}
    \includegraphics[width=1.\linewidth]{figures/smoothing_kernel.pdf}
    \oscaption{smoothing_kernel}{%
       Normalized intensity of a spherical harmonic expansion of 
        a Gaussian spot placed at zero degrees latitude and longitude.
        The expansion is in the quantity $\cos\Delta\theta$ where $\Delta\theta$ is the angular separation between the center of the spot and another point on the surface of the sphere, it is set to 5 degrees.
        The three insets show the spot profile at different strengths of the smoothing kernel
        where the leftmost spot has no smoothing ($\sigma_s=0$) and the rightmost spot is strongly smoothed. 
        The lines of different colors correspond to spot expansions up to certain order and the black line is the expansion at infinite order.
        The purpose of smoothing is to taper the higher order spherical harmonic coefficients.
        No smoothing results in significant "rings" where the intensity goes negative, even at high order.
        High levels of smoothing reduce the ringing but increase the spot size and eliminate differences between expansions above a certain order.
                \label{fig:smoothing_kernel}
    }
    \end{centering}
\end{figure}

Thus, there is a trade-off between smoothing and resolution of the maps. 
In principle we can always get rid of ringing by fitting sufficiently high order maps.
In practice, computing the analytic integrals computed by \textsf{starry} becomes computationally unstable above $l\approx 25$ so instead of going to very high order we apply some smoothing to mitigate the ringing.

\subsection{Fitting simulated ingress/egress light curves}
In Figure~\ref{fig:ingress_egress_simulated} we test a model with the smoothing filter 
by simulating data as in Figure~\ref{fig:schmixels_vs_harmonics}, except this time we simulate two light curves, an ingress light curve generated from real ephemeris of Io disappearing behind the limb of Jupiter and egress light curve generated when Io reappears.

\begin{figure}[h!]
    \begin{centering}
    \includegraphics[width=1.\linewidth]{figures/ingress_egress_simulated.pdf}
    \oscaption{ingress_egress_simulated}{%
        Maximum A Posteriori (MAP) estimate of an $l=25$ map (right panel from the top) produced by fitting a pair of light curves simulated from the true map which is shown on the left.
        The simulated light curves are shown in the third panel from the top, an ingress light curve on the left and an egress light curve on the right.
        The small circles above each of the light curves show the progression of the occultation over the inferred map.
        The precise location of the true spot is not recovered for either of the two models
        which is not surprising because a single light curve is not sufficient to break the degeneracies.
        \label{fig:ingress_egress_simulated}
    }
    \end{centering}
\end{figure}

\subsection{Fitting IRTF ingress/egress light curves}
\begin{figure}[h!]
    \begin{centering}
    \includegraphics[width=1.\linewidth]{figures/irtf_ingress_egress.pdf}
    \oscaption{irtf_ingress_egress}{%
Maximum A Posteriori (MAP) estimate of two $l=25$ maps produced by simultaneously fitting two IRTF light curves of an occultation of Io by Jupiter.
The light curve on the left corresponds to an ingress of Io 
        while the one on the right
        fitting a pair of light curves simulated from the true map which is shown on the left.
        The simulated light curves are shown in the third panel from the top, an ingress light curve on the left and an egress light curve on the right.
        The small circles above each of the light curves show the progression of the occultation over the inferred map.
        The precise location of the true spot is not recovered for either of the two models
        which is not surprising because a single light curve is not sufficient to break the degeneracies.
        \label{fig:irtf_ingress_egress}
    }
    \end{centering}
\end{figure}

\begin{figure}[h!]
    \begin{centering}
    \includegraphics[width=.5\linewidth]{figures/irtf_ingress_egress_loki.pdf}
    \oscaption{irtf_ingress_egress_loki}{%
        This is a plot of a pretty function. And at the end of this
        caption is a symbol with a link to the \emph{exact} script
        that generated it, hosted on \textsf{GitHub}.
        \label{fig:irtf_ingress_egress_loki}
    }
    \end{centering}
\end{figure}

>>>>

\begin{abstract}
Jupiter's moon Io is the most volcanically active body in the Solar System with hundreds of active volcanoes varying in intensity on different timescales.
Existing photometric observations in the near infrared taken during occultations by Jupiter and Galilean moons encode a wealth of information about its surface.
Io is thus an ideal testbed for models of time-variable exoplanet surfaces.
We built a generative model of Io's surface for each observation using Nonnegative Matrix Factorization.
The model is built on top of the code starry \href{https://rodluger.github.io/starry/}{\color{linkcolor}\faGithub} which enables fast analytic computation of occultation light curves, phase curves in both emitted and reflected light and also computes exact gradients with respect to all input parameters.
We use the inferred map to study the time variability of prominent volcanoes over a timescale of decades.
While we mostly focus on Io, the models are directly applicable to mapping exoplanets.
We discuss this application towards the end of the paper.\href{https://github.com/fbartolic/volcano}{\color{linkcolor}\faGithub}

@@

\subsection{Schmixels vs. spherical harmonics}
\label{ssec:schmixels_vs_harmonics}
To test inference with the static map model defined in Section~\ref{ssec:static} 
we use \textsf{starry} to generate a single simulated light curve of an occultation of Io by Jupiter from an $l=30$ map.
The simulated map consists of a spherical harmonic expansion of a bright spot with a Gaussian profile which we add using the built in \textsf{add\_spot} function in \textsf{starry}.
The expansion is in the quantity $\cos(\Delta\theta)$ where $\Delta\theta$ is the angular separation between the center of the spot and another point on the surface of the sphere. 
We place the spot at 13 degrees latitude and 51 degrees East longitude, set the diameter of the spot ($2\Delta\theta$) to 5 degrees and the amplitude to 55\% of the total luminosity of a spotless map.
We take ephemeris parameters (assumed to be fixed) from a real occultation and assume an observational cadence of 1s and errorbars equal to 2\% of the total flux, similar to high quality IRTF light curves.
We use this dataset to test the difference between fitting a model in the spherical harmonic basis (Eq.~\ref{eq:linear_model}) and in the schmixel basis (Eq.~\ref{eq:linear_model_pix})
by fitting for an $l=20$ map.
Since we are only fitting a single light curve, the inferred maps will be strongly prior dominated.

@@

The schmixel model with Half Gaussian priors produces a simpler map in which the unobserved hemisphere has much lower intensity but the arc is less sharply defined and more extended and as a result the predicted flux does not rise steeply enough compared to the light curve.
Finally, the Exponential prior on schmixels results in a map with the most spot like feature but it also has noticeable ringing.
This is somewhat surprising because based on Figure~\ref{fig:schmixels_to_pixels}, we wouldn't expect much difference between the Half Gaussian and the Exponential priors.
It seems that the heavier tail of the Exponential is nevertheless able to induce a sparser solution.
In the next section, we tackle the issue of ringing artefacts.

\subsection{Smoothing out spurious features}
\label{ssec:spurious_features}
To suppress ringing which results in regions of negative intensity surrounding a spot like feature as in Figure~\ref{fig:schmixels_vs_harmonics}, we apply a spatial smoothing filter to the spherical harmonics.
Mathematically, the filtering operation is a convolution between the map and some kernel function $B(\theta,\phi)$.
Assuming both the map and the kernel function are expanded in terms of spherical harmonics, the convolution operation is simply a multiplication between the two sets of spherical harmonic coefficients.
We choose a simple azimuthally symmetric Gaussian kernel function given by
\begin{equation}
    B(\theta)=\frac{1}{\sqrt{2 \pi \sigma_s^{2}}}\exp \left(-\theta^{2} / 2 \sigma_s^{2}\right)
\end{equation}
where $\sigma_s$ sets the characteristic scale of the smoothing.
This function can be expanded in terms of spherical harmonics as
\begin{equation}
    B(\theta)=\sum_{l=0}^{\infty}\left(\frac{2 l+1}{4 \pi}\right) B_{l} \,\mathcal{P}_{l}(\cos \theta)
\end{equation}
where $B_l$ are the spherical harmonic coefficients and $\mathcal{P}_l$ are the associated Legendre polynomials.
They depend only on $l$ because all nonzero $m$ modes vanish due to azimuthal symmetry.
For $\sigma_s\ll 1$ $B_l$ can be approximated as \citep{seon2007,white1995}
\begin{equation}
    B_l\simeq \exp\left[-\frac{1}{2}l(l+1)\sigma_s^2\right]
\end{equation}
The effect of this filter is to exponentially suppress features on scales smaller than $l\sim \sigma_s^{-1}$.

Figure~\ref{fig:smoothing_kernel} shows the effect of (Gaussian) smoothing on a spherical harmonic expansion of a spot with a Gaussian intensity profile.
We place the spot at zero degrees latitude and longitude and set the size of the spot $\Delta\theta$ to 5 degrees.
All three subplots show the exact profile of the spot (black line) and expansions up to three different orders $l$ (colored lines).
The subplot on the left shows the expansion with no smoothing ($\sigma_s=0$) in which case the symmetric ringing around the center of the spot is clearly visible even at relatively high order ($l=20$).
The middle subplot shows an intermediate level of smoothing where $\sigma_s=0.1$, meaning that all features on scales above $l\approx 10$ are exponentially suppressed.
The negative ringing is a lot less visible albeit at the cost of having a slightly larger spot because suppressing higher order harmonics necessarily means that we lose some ability to represent smaller scale features.
For $\sigma_s=0.2$ (right) there is practically no ringing, but the expansions at $l=10$, $l=15$ and $l=20$ result in the spot of the same size because all coefficients above $l=5$ are significantly suppressed.

\begin{figure}[h!]
    \begin{centering}
    \includegraphics[width=1.\linewidth]{figures/smoothing_kernel.pdf}
    \oscaption{smoothing_kernel}{%
       Normalized intensity of a spherical harmonic expansion of 
        a Gaussian spot placed at zero degrees latitude and longitude.
        The expansion is in the quantity $\cos\Delta\theta$ where $\Delta\theta=5^\circ$.
        The three plots show the spot profile for increasing values of the smoothing parameter 
    $\sigma_s=0$. 
        The colored lines correspond to spot expansions up to a certain order and the black line is the exact expansion.
        The purpose of smoothing is to taper higher order spherical harmonic coefficients in order to suppress ringing artefacts which result in negative intensities.
        It is apparent that no smoothing results in significant ringing, high levels of smoothing suppress the ringing completely but as a result they increase the spot size and eliminate differences between expansions above a certain order.
        Intermediate levels of smoothing provide a compromise beteween the two extremes.
               \label{fig:smoothing_kernel}
    }
    \end{centering}
\end{figure}

Thus, there is a trade-off between smoothing and being able to resolve smaller scale features
in the maps. 
In principle we can always get rid of ringing by fitting sufficiently high order maps.
In practice, computing the analytic integrals computed by \textsf{starry} becomes computationally unstable above $l\approx 25$ so instead of going to very high order we apply some smoothing to mitigate the ringing.
We find that setting $\sigma_s=2/l$ where $l$ is the order of expansion of the map provides the best results .

\subsection{Fitting simulated ingress/egress light curves}
In this section we test a model which includes the smoothing filter. 
We use the same map to generate the data as in Figure~\ref{fig:schmixels_vs_harmonics} except this time we simulate two light curves in order to break the degeneracy in the location of the spot, an ingress light curve generated from real ephemeris of Io disappearing behind the limb of Jupiter and egress light curve generated when Io reappears.
We generate the light curves from an $l=25$ map and fit a map of the same order. 
We apply a Gaussian smoothing filter to both the simulated map and the map we fit to the data and set the smoothing parameter to $\sigma_s=2/l=2/25$. 
We use an Exponential prior on the schmixels and set the scale parameter to 10.
The results are shown in Figure~\ref{fig:ingress_egress_simulated}.

\begin{figure}[h!]
    \begin{centering}
    \includegraphics[width=1.\linewidth]{figures/ingress_egress_simulated.pdf}
    \oscaption{ingress_egress_simulated}{%
Maximum A Posteriori (MAP) estimate of an $l=25$ map (right) produced by fitting a pair of simulated ingress and egress occultation light curves (bottom) which were generated from a $l=25$ map consisting of a single bright spot (left). 
       A Gaussian smoothing filter is applied to reduce the ringing artefacts.
       Small circles beneath each inferred map show the progression of the occultation.
        The model is able to recover the position of the true spot (marked with a grey cross) and there is noticeably less ringing than in Figure~\ref{fig:schmixels_vs_harmonics}.
        \label{fig:ingress_egress_simulated}
    }
    \end{centering}
\end{figure}

The inferred map looks significantly better than the one in Figure~\ref{fig:schmixels_vs_harmonics}. 
With two occultation light curves the model is able to break the degeneracy in the location of the spot and thanks to the smoothing filter there is almost no visible ringing around the spot in the inferred map.
Some residuals in the data are still present (WHY).
In the next section, we fit a pair of light curves from the IRTF dataset described in Section~\ref{sec:data}.

\subsection{Fitting IRTF ingress/egress light curves}

\begin{figure}[h!]
    \begin{centering}
    \includegraphics[width=1.\linewidth]{figures/irtf_ingress_egress.pdf}
    \oscaption{irtf_ingress_egress}{%
        Maximum A Posteriori (MAP) estimate of $l=25$ maps (top) produced by fitting a pair of IRTF ingress (left) and egress (right) occultation light curves of Io by Jupiter from 1998.
       The small circles beneath each inferred map show the progression of the occultation.
        Both light curves are assumed to be generated from the same map except for a difference in the overall amplitude.
        We allow for multiplicative rescaling of the error bars and an additive flux offset for each light curve.
        The model places a single bright spot to explain most of the variance in the data.
        The visible patterns in the residuals are likely due to additional hot spots and the limited resolution of our maps.
        \label{fig:irtf_ingress_egress}
    }
    \end{centering}
\end{figure}

\begin{figure}[h!]
    \begin{centering}
    \includegraphics[width=.5\linewidth]{figures/irtf_ingress_egress_loki.pdf}
    \oscaption{irtf_ingress_egress_loki}{%
        Hot spot from the inferred maps shown in Figure~\ref{fig:irtf_ingress_egress} overlaid on top of the U.S. Geological Survey's map of the surface of Io (CITE) which was
        constructed from Galileo observations.
        The location of the hot spots matches the location of Loki Patera, the largest active volcano on the surface of Io. 
        \label{fig:irtf_ingress_egress_loki}
    }
    \end{centering}
\end{figure}
